Calibrated, Parametrized Learning

### 2014-11-12 tex/parametrized-learning.tex
progress on note after UCI seminar

\begin{itemize}
 \item $x$: a vector of observed quantities for each event (energy, momenta, angles of particles)
 \item $\theta$: parameters of a statistical model
\item $f_c(x| \theta)$:  probability density function (statistical model) for $x$ for category $c\in\{0,1\}$.
\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta)$: real-valued score from a parametrized machine learning classification algorithm (or any map $s: X\otimes \Theta \to\mathbb{R}$)
\item $f_c( s | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x)$
\item $f_c( s(\cdot; \theta) | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x;\theta)$

@@

These steps lead to a subsequent statistical analysis where one observes in data $\{x_e\}$, where $e$ is an event index running from $1$ to $n$. For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution. In particular, one forms the statistical model
\begin{equation}\label{eq:typicalML}
p( \{ x_e \} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e) \,|\, \theta) \,\right] \; ,
\end{equation}
where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known.} Typically, we are interested in inference on $\mu$ and $\theta$ are nuisance parameters; though, sometimes $\theta$ may include some components that we are also wish to infer (like the mass of a new particle that affects the distribution $x$ for the signal events).

@@

The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $s(x; \theta)$. Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\theta$ and then includes the corresponding value of $\theta$ as an input to the classifier. Thus $\{x_e,c_e\} \to \{x_e,\theta_e, c_e\}$ leading to a parametrized learner $s(x)\to s(x;\theta)$. This leads to a complication: one does not know the value of $\theta$ to use when evaluating the parametrized classifier $s(x;\theta)$.  However, when performing statistical inference one will evaluate the likelihood of the data at a given value of $\theta$, so this is a natural choice.  Thus one arrives at the generalization of Eq.~\ref{eq:typicalML}
\begin{equation}\label{eq:parametrizedML}
p( \{ x_e \} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e;\theta) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e;\theta) \,|\, \theta) \,\right] \; .
\end{equation}

@@

>>>>

\begin{itemize}
 \item $x$: a vector of observed quantities for each event (energy, momenta, angles of particles)
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$
 \item $\theta$: parameters of a statistical model
\item $f_c(x| \theta)$:  probability density function (statistical model) for $x$ for category $c\in\{0,1\}$.
\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta)$: real-valued score from a parametrized machine learning classification algorithm (or any map $s: X\otimes \Theta \to\mathbb{R}$)
\item $f_c( s | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x)$
\item $f_c( s(\cdot; \theta) | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x;\theta)$

@@

These steps lead to a subsequent statistical analysis where one observes in data $\{x_e\}$, where $e$ is an event index running from $1$ to $n$. For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution. In particular, one forms the statistical model
\begin{equation}\label{eq:typicalML}
f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e) \,|\, \theta) \,\right] \; ,
\end{equation}
where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known.} Typically, we are interested in inference on $\mu$ and $\theta$ are nuisance parameters; though, sometimes $\theta$ may include some components that we are also wish to infer (like the mass of a new particle that affects the distribution $x$ for the signal events).

@@

The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $s(x; \theta)$. Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\theta$ and then includes the corresponding value of $\theta$ as an input to the classifier. Thus $\{x_e,c_e\} \to \{x_e,\theta_e, c_e\}$ leading to a parametrized learner $s(x)\to s(x;\theta)$. This leads to a complication: one does not know the value of $\theta$ to use when evaluating the parametrized classifier $s(x;\theta)$.  However, when performing statistical inference one will evaluate the likelihood of the data at a given value of $\theta$, so this is a natural choice.  Thus one arrives at the generalization of Eq.~\ref{eq:typicalML}
\begin{equation}\label{eq:parametrizedML}
f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e;\theta) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e;\theta) \,|\, \theta) \,\right] \; .
\end{equation}

@@

\subsection{A different starting point}

Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $f_c(x|\theta)$ and a well defined objective.

\subsubsection{Simple Hypothesis Testing}

Consider the situation of simple hypothesis testing between the background-only $\mu=0$ and signal+background hypothesis $\mu=\mu_0=\nu_s/(\nu_s+\nu_b)$ for $\theta=\theta_0$.  Let us assume that not only can we use the $f_c(x|\theta_0)$ as a generative model, but that we can evaluate the probability density at any point $x$.  The statistical model is given by
\begin{equation}\label{eq:NP}
f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( x_e \, |\,  \theta)  + (1-\mu)\, f_0( x_e \,|\, \theta) \,\right] \; .
\end{equation}
The objective is to find a  test statistic $T(D)$ (a map from the space of the data to $\mathbb{R}$)  
and a threshold $k_\alpha$ that minimizes the rate of type-II error (i.e. $P(T(D) < k_\alpha | \mu=\mu_0, \theta_0) = \beta$ under the constraint of a fixed rate of type-I error ( i.e.  $P(T(D) > k_\alpha | \mu=0, \theta_0) = \alpha$ ).
The Neyman-Pearson lemma states that the optimal test statistic is the likelihood ratio, or equivalently, the log-likelihood ratio 
\begin{equation}
T =  \log \frac{f(D | \mu=\mu_0, \theta_0)}{f(D | \mu=0, \theta_0)}
\end{equation}
When the $x_e$ are i.i.d., then the log-likelihood ratio expands into a sum
\begin{equation}
T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(x_e | \theta_0) }{ f_0(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
\end{equation}
where $c_1=\mu/(1-\mu)$ and $c_2=\log(1-\mu)$ are constants for the simple hypothesis test.

Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$. A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.

The important part of the per-event function $q(x)$ is that it defines contours in the feature space $x$. These contours are also equivalent to the function $f_1(x|\theta_0)/f_0(x|\theta_0)$, the signal-to-background ratio. If we can find a machine learner $s(x)$ that is a monotonic function of $f_1(x|\theta_0)/f_0(x|\theta_0)$ it will share the same contours. Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.

Postulate:
\[
T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(s(x_e) | \theta_0) }{ f_0(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
\]
leads to an equivalent test, where $s_e \equiv s(x_e)$
\[
f_c(s) = \int dx \, \delta(s-s(x)) \, f_c(x)  \,  |\partial s / \partial x|^{-1}
\]

Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$.

Proof:

First show $q(x_e) = q(s(x_e))$.
\[
\frac {f_1(s(x) | \theta_0) }{ f_0(s(x) | \theta_0) } = \frac { \int dx \delta(s-s(x)) \, f_1(x_e | \theta_0) J }{ \int dx \delta(s-s(x)) \, f_0(x_e | \theta_0) J } = \frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
\]

Second, show density is the same

\[
f_c(q_x) = \int dx \delta(q_x-q_x(x)) f_c(x) / |\partial q_x / \partial x|
\]

\[
f_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, f_c(x) \, / |\partial q_s / \partial x|
\]

\[
f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
\]

This is the test statistic that one would naturally get from constructing the likelihood function for the 1-d distributions of the output of the machine learner
\begin{equation}\label{eq:NP}
f( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s_e \, |\,  \theta)  + (1-\mu)\, f_0( s_e \,|\, \theta) \,\right] \; .
\end{equation}
Thus this motivates Eq.~\ref{eq:typicalML} in the more general case.

Question:
Where does this fit in:
\[
T'' = \sum_{e=1}^n \underbrace{ \log \left[ 1+c_1 s(x_e) \right] }_{q'(s_e)} \;,
\]

Usefulness:
If we can readily compute $f_c(x|\theta_0)$, then there would be no need for machine learning algorithms. However,  these densities are often difficult to compute though one can readily use them as a generative model to produce samples. In that case, one might wish to use a machine learning algorithm to find $s(x) \approx f_1(x|\theta_0) / f_0(x|\theta_0)$.

Conclusion, while it is possible to start directly with the Neyman-Pearson loss function of type-II error under the constraint of fixed size and try to find a per-experiment learner $T(D)$, it is more convenient to cast the problem in terms of learning a per-event function $s(x)$, and then calibrate it to be a linear function of $q(x)$.
